DRAFT: The algebraic hierarchy in type theory using type classes

### 2010-03-19 jar.tex
Fix

\documentclass[a4paper,10pt, runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{ucs}
\usepackage{amsmath}
% \usepackage[utf8x,utf8]{inputenc}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%include coq2tex.tex
\arrayhs%i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\begin{document}
\title{DRAFT: The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We propose generic design patterns to organize algebraic structures using type classes inside the
Coq system. We support multiple inheritance, sharing of notations and theories, and automated
structure inference. A library for many-sorted universal algebra abstracts from the specific
algebraic structures and aids the organization of the library. Category theory provides a further
useful abstraction. We provide abstract
interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
fractions. Similarly, we provide an interface for polynomials over a ring $R$ using $R$-algebras.
We show that the standard implementations of these are in fact instances of these type classes.
We report on the efficiency penality of these abstractions.

@@

\section{Preliminaries}
\subsection{Coq type theory}
The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}.\marginpar{More}
This type theory lacks quotient types, as it would make type checking undecidable. Instead it
supports setoids (aka Bishop sets), a pair of a type and an equivalence
relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support has been added to the
system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called |Proper|.
As we will see in Section~\ref{univ} below this functionality can also be used to work with
quotient algebras.

Coq has two kinds of sorts |Prop| for propostions and |Type| for data types. Coq
is scheduled to support proof irrelevance: every |p,q:P:Prop| are
convertible in the future; see section~\ref{PI}.

@@

\section{To bunde or not to bundle? (Or: What to bundle?)}

How should we define reflexive, transitive, and symmetric relations? In a first attempt, we
might define:
%format → = "\rightarrow"
%format Π = "\Pi"
\begin{code}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: rr_carrier → rr_carrier → Prop
    ; rr_proof: Π x, rr_rel x x }.
\end{code}
and similar for transitive and symmetric relations.

With this approach, we run into problems when we try to define |EquivalenceRelation| in terms of
|ReflexiveRelation| and its two siblings, because when posit three objects of type
|ReflexiveRelation|, |TransitiveRelation|, and |SymmetricRelation|, respectively, we have to
add equalities expressing that they talk about the same carrier and relation, and these equalities
are a nightmare to work with (because the three carriers and relations won't be convertible).
Technically, this leads to the research in manifest fields, the \emph{with} construction, to which
we return in Section~\ref{manifest}.

As a first step we would define |ReflexiveRelation| as:
\begin{code}
  Record ReflexiveRelation (carrier: Type): Type :=
    { rr_rel: rr_carrier → rr_carrier → Prop
    ; rr_proof: Π x, rr_rel x x }.
\end{code}

@@

A natural solution is to pull the relation out of |ReflexiveRelation| (and its siblings) as well:
\begin{code}
  Record ReflexiveRelation (carrier: Type)
                           (rel: carrier → carrier → Prop): Prop :=
    { rr_proof: Π x, rr_rel x x }.
\end{code}
or equivalently:
\begin{code}
  Definition ReflexiveRelation (carrier: Type)
                               (rel: carrier → carrier → Prop): Prop :=
    Π x, rr_rel x x.
\end{code}
This is how reflexivity is defined in the standard library, and this is entirely sensible. The
change from |Type| to |Prop| reflects the fact that |ReflexiveRelation| is now strictly a property,
proofs of which can be nicely kept opaque;%
\footnote{Coq distinguishes between transparent and opague definitions. The former can be unfolded,
whereas this is impossible for the latter.}
we will return to this issue in Section~\ref{PI}.
We can now define |EquivalenceRelation| without awkward equalities:
\begin{code}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: er_carrier → er_carrier → Prop
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

@@

\begin{code}
Record EquivalenceRelation (er_carrier: Type)
                             (er_rel: er_carrier → er_carrier → Prop): Prop :=
  { er_refl: ReflexiveRelation er_carrier er_rel
  ; er_sym: SymmetricRelation er_carrier er_rel
  ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

@@

At the risk of overstressing the point, let us look at how the same problem and solution keep
appearing in composite structures. Suppose we use a definition of |Monoid| that bundles the
relation. Then if we proceed to define |SemiRing| as a double |Monoid|, we face a dilemma:
\begin{code}
  Record SemiRing A: Type :=
    { mult_monoid: Monoid A 
    ; plus_monoid: Monoid A 
    ; plus_comm: Commutative (mon_op plus_monoid)
    ; mult_comm: Commutative (mon_op mult_monoid)
    ; mult_0_l: Π x: A, ?
\end{code}
Which equality do we use in the expression ?, |mult_monoid|'s equality or |plus_monoid|'s?
Again, for this definition to be usable at all, we would need to add an equality expressing that the
two monoids contain the same equivalence relation. And again, the proper solution is to change
Monoid and unbundle the relation. But here, we have a choice as to how much we unbundle exactly.
(For the moment we ignore the question of whether the unit and binary operator should be unbundled
as well, focusing only on the equivalence relation and its equivalence proof.)

@@

Let us now stop and think to what extent this approach suffers from all the problems commonly
associated with it. In particular, let us imagine what happens to our terms and contexts when we
want to talk about nested structures such as polynomials over the ring of matrices over a
given ring; see Section~\ref{interfaces}.

@@

@@

We have developed basic universal algebra~\cite{meinke1993universal} upto the first homomorphism
theorem.~\marginpar{Description of the work.}
Discuss: the duplication of code which may have been avoided by universe
polymorphism for |Fixpoint|s.\marginpar{universe polymorphism}
%perhaps the soundness theorem for equational logic?

In order to prove the first homophism theorem, we define a congruence.
\begin{code}
Class Congruence: Prop :=
    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
    ; congruence_quotient:> @Algebra et v e _
    }.
\end{code}
Alternatively, we could have state the compatibility of the relation |e| by stating that the
relation as a set of pairs is a subalgebra in the product algebra. We prove that both approaches
are equivalent.

A congruence relation is coarser than the setoid relation, as indicated by the use of |Proper|
above. Since our theory is founded on a congruence, the initial setoid equality, the development of 
congruences and quotients benefits from this.\marginpar{how}

We define the forgetful functor from the category of $\tau$-algebras to the category of sets.
We define the term algebra consisting of closed terms, terms with the empty type as set of
variables. The term algebra is the free $\tau$-algebra and hence\marginpar{?}
initial. Categorically: we have defined a left-adjoint to the forgetful functor. A formal proof
connects the categorical and the algebraic definition.
One may expect that this term algebra could be useful to
\emph{define}, say, the natural numbers as the term algebra for the theory of semiring. However, it
seems difficult to prove decidability of equality, which requires a normalization procedure.

@@

%format ◎ = "\circ"
%interfaces/abstract_algebra.v
\begin{code}
Class Category O `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
  `{!CatId O} `{!CatComp O}: Prop :=
  { arrow_equiv:> Π x y: O, Setoid (x ⟶ y)
  ; comp_proper:> Π x y z,
      Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
  ; comp_assoc (w x y z: O) (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
      c ◎ (b ◎ a) = (c ◎ b) ◎ a
  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
\end{code}
This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
Similarly, we have equality on natural transformations, but not on functors.

%theory/categories.v
Natural transformations are formalized as:
\begin{code}
  Context `{Category C} `{Category D}
  `{!Functor (F: C → D) Fa} `{!Functor (G: C → D) Ga}.

  Class NaturalTransformation (η: Π c, F c ⟶ G c): Prop :=
    natural: Π (x y: C) (f: x ⟶ y), η y ◎ fmap F f = fmap G f ◎ η x.
\end{code}

@@

@@

@@

\section{Type class quoting}\label{quote}
Unification hints~\cite{Hints} allow one to semi-automatically construct a quote
function\footnote{Gonthier provides similar functionality using canonical structures}. However,
this feature is absent from Coq. Fortunately, type classes provide similar functionality as
we will now show.

@@

For our methodology to work we have to be able to transform the occuring record into a telescopic
subset $\{ A \mid  B\}$. We moreover conjecture that the terms in |B| only depend on the
variables in |A|, but not in |B| itself. However, we do not need this.

@@

We conclude that unification hints, canonical structures and type classes all appear to have
similar expressive powers. Canonical structures, being tied to the unifier, are more robust, but
seem to require more inguinity. Type classes are easier to use. Unification hints, may be a
good generalization, but are absent from Coq. We encourage their inclusion to the
Coq system.

>>>>

\documentclass[a4paper,10pt,runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{amsmath}
\usepackage[utf8x]{inputenx}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%include polycode.fmt
%include coq2tex.tex
\arrayhs%i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\newcommand{\textPi}{\ensuremath\Pi}
\newcommand{\textSigma}{\ensuremath\Sigma}
\newcommand{\texttau}{\ensuremath\tau}
% \newcommand{\textforall}{\ensuremath\forall}
\begin{document}
\title{DRAFT: The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We\footnote{\today} propose generic design patterns to organize algebraic structures using
type classes inside the
Coq system. We support multiple inheritance, sharing of notations and theories, and automated
structure inference. A library for many-sorted universal algebra abstracts from the specific
algebraic structures and aids the organization of the library. Category theory provides a further
useful abstraction. We provide abstract
interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
fractions. Similarly, we provide an interface for polynomials over a ring $R$ using $R$-algebras.
We show that the standard implementations of these are in fact instances of these type classes.
We report on the efficiency penality of these abstractions.

@@

\section{Preliminaries}
\subsection{Coq type theory}
The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types.
We will not elaborate on this, but instead refer to~\cite{Coq,BC04}.

The Coq type theory lacks quotient
types, as it would make type checking undecidable. Instead it
supports setoids (aka Bishop sets), a pair of a type and an equivalence
relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support for setoid has been added to
the system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called
|Proper|.
As we will see in Section~\ref{univ} below this functionality can also be used to work with
quotient algebras.

Coq has two kinds of sorts |Prop| for propostions and |Type| for data types. Coq
is scheduled to support proof irrelevance, every |p,q:P:Prop| are
convertible, in the future; see section~\ref{PI}.

@@

\section{To bunde or not to bundle? (Or: What to bundle?)}
We provide a rationale for our approach. We begin gently.

How should we define reflexive, transitive, and symmetric relations? In a first attempt, we
might define:
%format → = "\rightarrow"
%format Π = "\Pi"
\begin{code}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{code}
and similar for transitive and symmetric relations.

With this approach, we run into problems when defining |EquivalenceRelation| in terms of
|ReflexiveRelation| and its two siblings, because when posit three objects of type
|ReflexiveRelation|, |TransitiveRelation|, and |SymmetricRelation|, respectively, we have to
add equalities expressing that they talk about the same carrier and relation. These equalities
cause difficulties, because the three carriers and relations won't be convertible.
Technically, this leads to the research in manifest fields, i.e.\ the \emph{with} construction, to
which
we return in Section~\ref{manifest}.

As a first step we would define |ReflexiveRelation| as:
\begin{code}
  Record ReflexiveRelation (carrier: Type): Type :=
    { rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{code}

@@

A natural solution is to pull the relation out of |ReflexiveRelation| (and its siblings) as well:
\begin{code}
  Record ReflexiveRelation (carrier: Type)
                           (rel: relation carrier): Prop :=
    { rr_proof: Π x, rr_rel x x }.
\end{code}
or equivalently:
\begin{code}
  Definition ReflexiveRelation (carrier: Type)
                               (rel: relation carrier): Prop :=
    Π x, rr_rel x x.
\end{code}
This is how reflexivity is defined in the standard library, and this is entirely sensible. The
change from |Type| to |Prop| reflects the fact that |ReflexiveRelation| is now strictly a property,
proofs of which can be nicely kept opaque;%
\footnote{Coq distinguishes between transparent and opague definitions. The former can be unfolded,
whereas this is impossible for the latter. Since we never need to distinguish between two proofs of
a propositions, Propositions can be Opague.}
we will return to this issue in Section~\ref{PI}.
We can now define |EquivalenceRelation| without awkward equalities:
\begin{code}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: relation er_carrier
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

@@

\begin{code}
Record EquivalenceRelation (er_carrier: Type)
                             (er_rel: relation er_carrier): Prop :=
  { er_refl: ReflexiveRelation er_carrier er_rel
  ; er_sym: SymmetricRelation er_carrier er_rel
  ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

@@

At the risk of overstressing the point, let us look at how the same problem and solution keep
appearing in composite structures. Suppose we use a definition of |Monoid| that bundles the
relation. Then if we proceed to define |SemiRing| as a double |Monoid|, we face a dilemma:
\begin{code}
  Record SemiRing A: Type :=
    { mult_monoid: Monoid A 
    ; plus_monoid: Monoid A 
    ; plus_comm: Commutative (mon_op plus_monoid)
    ; mult_comm: Commutative (mon_op mult_monoid)
    ; mult_0_l: Π x: A, H
\end{code}
Do we use |mult_monoid|'s equality or |plus_monoid|'s equality in |H|?
Again, for this definition to be usable at all, we would need to add an equality expressing that the
two monoids contain the same equivalence relation. And again, the proper solution is to change
Monoid and unbundle the relation. But here, we have a choice as to how much we unbundle exactly.
(For the moment we ignore the question of whether the unit and binary operator should be unbundled
as well, focusing only on the equivalence relation and its equivalence proof.)

@@

@@

Let us now stop and think to what extent this approach suffers from all the problems commonly
associated with it. In particular, let us imagine what happens to our terms and contexts when we
want to talk about nested structures such as polynomials over the ring. For a concrete
representation of the polynomials the context will just contain the context for an abstract ring.
For abstract reasoning about polynomials the context will grow with abstract operations on
polynomials. Consequently, the context will grow linearly, as opposed to exponentially.

@@

We have developed basic multi-sorted universal algebra~\cite{meinke1993universal} upto the first
homomorphism theorem. More precisely, we provide the following:
\begin{itemize}
 \item Definition of a signature Σ. Definition of a type (Σ,E). Σ-algebras and
(Σ,$E$)-algebras
 \item A forgetful functor from (Σ,$E$)-algebras via Σ-algebras to Sets
 \item Products and subalgebras
 \item The term algebra for a type $\tau$. This is an initial object in the category of
$\tau$-algebras.
\end{itemize}
%perhaps the soundness theorem for equational logic?

Coq provides different sorts. To simplify the discussion, we assume that there are only two sorts:
|Prop| and |Type|. It occassionally happens that we want to prove a similar theorem for both sorts.
To avoid code duplication, Coq provides universe polymorphism for inductive types.
Unfortunately,
definitions and fixpoints are not (yet?) polymorphic. To avoid code duplication we
change some Fixpoints into
Inductives. Concretely, the image $Im f:=\{b:B \mid \exists a. b=(f
a)\}$ is informative (in |Type|) since we need the witness in the proof of the first homomorphism
theorem.\marginpar{Does |relation| as a definition cause universe inconsistencies?}
% theory/ua_subalgebraT.v
% find . -iname "*.v" |xargs grep "niverse"

In order to prove the first homophism theorem, we define a congruence.
\begin{code}
Class Congruence: Prop :=
    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
    ; congruence_quotient:> @Algebra et v e _
    }.
\end{code}
Alternatively, we could have stated the compatibility of the relation |e| by stating that the
relation as a set of pairs is a subalgebra in the product algebra. We prove that both approaches
are equivalent. The former definition seems more natural in type theory. In set theory, a quotient
is defined as a collection of equivalence classes. In type theory, a quotient is more naturally
defined by changing the equality, while leaving the carrier unchanged.%
\footnote{In Bishop's words~\cite[p.12]{Bishop/Bridges:1985}: The axiom of choice is used to extract
elements from equivalence classes where they should never have been put in the first place.}
 In this way we profit from
the infrastructure for setoids: A congruence relation is coarser than the setoid relation, as
indicated by the use of |Proper| above.

%varieties/closed_terms.v
We define the forgetful functor from the category of $\tau$-algebras to the category of sets.
We define the term algebra consisting of closed terms, i.e.\ terms with the empty type as set of
variables. The term algebra is the initial $\tau$-algebra. Categorically: we have defined a
left-adjoint to the forgetful functor.\marginpar{Did we proof this?}
A formal proof connects the categorical and the algebraic definition.
One may expect that this term algebra could be useful to
\emph{define}, say, the natural numbers as the term algebra for the theory of semiring. However, it
seems difficult to prove decidability of equality, which requires a normalization procedure.

@@

\newcommand{\rightarrowtext}{\ensuremath{\rightarrow}}
\newcommand{\textocirc}{\ensuremath\circ}
%interfaces/abstract_algebra.v

CODE MISSING.
% \begin{code}
% Class Category O `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
%   `{!CatId O} `{!CatComp O}: Prop :=
%   { arrow_equiv:> Π x y: O, Setoid (x ⟶ y)
%   ; comp_proper:> Π x y z,
%       Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
%   ; comp_assoc (w x y z: O) (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
%       c ◎ (b ◎ a) = (c ◎ b) ◎ a
%   ; id_l `(a: x ⟶ y): cat_id ◎ a = a
%   ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
% \end{code}
This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
Similarly, we have equality on natural transformations, but not on functors.

%theory/categories.v
Natural transformations are formalized as:
CODE MISSING.
% \begin{code}
%   Context `{Category C} `{Category D}
%   `{!Functor (F: C → D) Fa} `{!Functor (G: C → D) Ga}.
% 
%   Class NaturalTransformation (η: Π c, F c ⟶ G c): Prop :=
%     natural: Π (x y: C) (f: x ⟶ y), η y ◎ fmap F f = fmap G f ◎ η x.
% \end{code}

@@

Universe polymorphism allows us to define the category of categories. However, we need to avoid
making |Relation| a |Definition|, since |Definitions| are not universe polymorphic.

@@

Similarly, we use type classes to flexibly combine two representations of a dense set of the
reals.
Let |Q| be the rationals and |D| be the dyadic rationals, i.e.\ rationals where the denominator is
a power of 2. Let |inj:D>->Q| be a coercion. Then we want to define heterogeneous multiplication
as follows:

@@

\section{Type class quoting}\label{quote}
Unification hints~\cite{Hints} allow one to semi-automatically construct a quote
function.\footnote{Gonthier provides similar functionality using canonical structures.} However,
this feature is absent from Coq. Fortunately, type classes provide similar functionality as
we will now show.

@@

For our methodology to work we have to be able to transform the occuring record into a telescopic
subset $\{ A \mid  B\}$. We moreover conjecture that the terms in |B| only depend on the
variables in |A|, but not in |B| itself. However, we do not need this.\\

We observed that we were naturally lead to the following methodology: Let φ be a statement in type
theory, i.e.\ an alternation of Πs and Σs. By the axiom of choice, we first Skolemize to $Σ h. Πa.P$
and then transfer the variable $h$ to the parameters. For example, consider surjectivity. This is
skolemized to a surjection with a right inverse, i.e.\ a split epi. We then move the right inverse
to the parameters.

It is interesting to note that classical logic is conservative for purely universal statements.

The addition of implicit Sigma types to Coq would change many things. Perhaps it would even allow us
to pack the proofs together with the operations, but it is too early to tell.

@@

We conclude that unification hints, canonical structures and type classes all appear to have
similar expressive powers. Canonical structures, being tied to the unifier, are more robust, but
seem to require more inguinity. Type classes are easier to use. Unification hints, may be a
good generalization, but are absent from Coq. We encourage their inclusion to the
Coq system.\marginpar{for instance what about control in
proof/instance search?\\ or what about its efficiency?}
